Some examples in extended logic counting classes

### 2016-12-14 LICS/appendix.tex
avance en saluja

For the second condition, note that for each function $f\in \nfpspace$, $\vert f(\A) \vert$ is polynomial on $\vert A \vert$. Thus, there is a $\pfp$ formula $\Phi$ that models the string $f(\A)$ by the bit. The rest of the proof is analogous to Theorem \ref{captfp}.

@@

\textbf{Proof of Theorem \ref{saluja-eq}}

>>>>

For the second condition, note that for each function $f\in \nfpspace$, $\vert f(\A) \vert$ is polynomial on $\vert A \vert$. Thus, there is a $\pfp$ formula $\Phi$ that models the string $f(\A)$ bitwise. The rest of the proof is analogous to Theorem \ref{captfp}.

@@

\textbf{Proof of Theorem \ref{one-sigma-zero}}

Towards a contradiction, suppose that there exists a constant function $f(\A) = s > 0$ which is expressible in $\E{0}$. This is, there is a formula $\psi(x,X)$ such that $f_{\psi(x,X)}(\A) = \vert \{ \langle e,P\rangle \mid \A\models\psi(e,P) \} \vert = s$ for each $\A\in\ostr$. It can be seen that if $X$ contains more than zero variables, then the function $f_{\psi(x,X)}$ grows exponentially. We have now that $\psi(x)$ contains only first order variables. Consider some $\A \in \ostr$ of size $n$ and let $a_1,\ldots,a_s\in \A$ be the tuples that satisfy $\A \models \psi(a_i)$. Now consider a structure $\A'$ that is obtained by duplicating $\A$. This is, $\A'$ has size $2n$, its first half is isomorphic to its second half, and also to $\A$. We have that there is at least $2s$ tuples $a_i,b_i\in \A'$, $s$ on each half, such that $\A'\models\psi(a_i)$ and $\A'\models\psi(b_i)$, which leads to a contradiction.

\textbf{Proof of Theorem \ref{mult-sigma-zero}}

Let $\alpha \in \sqso(\loge{0})$ with binary product. This is, $\alpha$ is defined by the grammar $\alpha := \sa{x}\varphi \ \mid \ \sa{X}\varphi \ \mid \ \sa{x}\alpha \ \mid \ \sa{X}\alpha \ \mid \ (\alpha\mult\alpha)$, where $\varphi \in \loge{0}$. For convenience, we add the expressions $\sa{x}(\varphi\mult\varphi)$ and $\sa{X}(\varphi\mult\varphi)$ to the grammar. Let $k$ be the number of algebraic operators in $\alpha$. We will show by induction on $k$ that $\alpha$ is expressible without multiplication. If $k = 1$, then the possible cases are $\alpha = \sa{x}\varphi$, where $\varphi\in\loge{0}$, which is already product-free. If $k = 2$, then the only that is not two nested summations is $\alpha = \sa{x}(\varphi\mult\psi)$, which can be expressed as $\alpha = \sa{x}(\varphi\wedge\psi)$. Suppose that the hypothesis holds for $k = 1$ through $n-1$. Let $\alpha$ have $n$ algebraic operators. If $\alpha = \sa{x}\beta$ or $\alpha = \sa{X}\beta$ where $\beta$ is an algebraic expression, then $\beta$ has $n-1$ algebraic operators and by hypothesis it can be expressed as product-free. If $\alpha = (\beta \mult \gamma)$ there are two cases. If at least one between $\beta$ and $\gamma$ has a product, then by hypothesis it can be expressed as product-free. Moreover, the new expression for $\alpha$ has strictly less than $n$ algebraic operators so by hypothesis it can be expressed as product-free. If there are no products in $\beta$ or $\gamma$, then let $\beta = \sa{x}\delta(x)$. We have that $\alpha$ can be expressed as $\sa{y}(\delta(y)\mult\gamma)$, where $y$ is a first-order variable that does not appear in $\beta$ or $\gamma$. Since $\delta(y)\mult\gamma$ has $n-1$ algebraic operators, it can be expressed as product-free, and then $\alpha$ can be expressed as product-free.

\textbf{Proof of Theorem \ref{fo-prod-sigma-zero}}

The resulting grammar is $\alpha := \sa{x}\varphi \ \mid \ \sa{X} \varphi \ \mid \ \sa{x}\alpha \ \mid \ \sa{X}\alpha \ \mid \pa{x}\alpha$. Consider a more restricted grammar that replaces $\pa{x}\alpha$ by $\pa{x}\sa{y}\varphi$. Let $\alpha = \pa{x}\sa{y}(x = y)$. This formula produces a constant function equal to 1, which as shown in Theorem \ref{one-sigma-zero} is not expressible in $\E{0}$.

### 2016-12-14 LICS/saluja.tex
avance en saluja

Saluja et. al \cite{DBLP:journals/jcss/SalujaST95} define a family of counting classes $\#\cL$ for each fragment $\cL$ of $\fo$. For a formula $\varphi(x,X)$, the function $f_{\varphi(x,X)}$ is defined as
\[
f_{\varphi(x,X)}(\A) = \vert \{\langle e,P\rangle\mid \A\models\varphi(e,P)\}\vert.
\]
for each $\A\in\str$. A function $f:\Sigma^*\to\nat$ is in $\#\cL$ if there exists an $\cL$ formula $\varphi(x,X)$ such that $f = f_{\varphi(x,X)}$.

\begin{theorem} \label{saluja-eq}
	$\eqso(\cL) = \#\cL$ for each fragment $\L$ of $\fo$.
\end{theorem}

For every logic $\cL$, we define an $\cL$-extended quantifier-free (QF) formula as follows:
\begin{eqnarray*}
	\varphi &::=& \alpha, \alpha \text{ is an $\cL$-formula} \ \mid \\
	&& X_i(x_1,\dots,x_{a_i}), i\in\N \ \mid \ \\
	&& (\neg \varphi) \ \mid \ (\varphi \wedge \varphi) \ \mid \ (\varphi \vee \varphi).
\end{eqnarray*}

>>>>

Saluja et. al \cite{DBLP:journals/jcss/SalujaST95} define a family of counting classes $\#\LL$ for each fragment $\LL$ of $\fo$. For a formula $\varphi(x,X)$, the function $f_{\varphi(x,X)}$ is defined as
\[
f_{\varphi(x,X)}(\A) = \vert \{\langle e,P\rangle\mid \A\models\varphi(e,P)\}\vert.
\]
for each $\A\in\str$. A function $f:\Sigma^*\to\nat$ is in $\#\LL$ if there exists an $\LL$ formula $\varphi(x,X)$ such that $f = f_{\varphi(x,X)}$. Note that for each fragment $\LL$ of $\fo$, we have that $\#\LL \subseteq \eqso(\LL)$. In fact we need a more restricted fragment of $\eqso$. Let $\sqso(\LL)$ be a fragment of $\qso$ defined by the following grammar:
\[
\alpha := \sa{x}\varphi \ \mid \ \sa{X} \varphi \ \mid \ \sa{x}\alpha \ \mid \ \sa{X}\alpha,
\]
where $\varphi$ is an $\LL$-formula. It can be seen that $\sqso(\LL) = \#\LL$. We are interested in, for each $\fo$-fragment $\LL$, the biggest fragment of $\eqso$ that is contained in $\#\LL$.

Let $\LL = \loge{0}$:
\begin{theorem} \label{one-sigma-zero}
	Positive constant functions are not expressible in $\E{0}$
\end{theorem}
\begin{corollary}
	If a fragment of $\eqso(\loge{0})$ is contained in $\E{0}$, its grammar does not allow sole constants.
\end{corollary}
\begin{proposition}
	Sum is not expressible in $\E{0}$
\end{proposition}
\begin{theorem} \label{mult-sigma-zero}
	$\sqso(\loge{0})$ with binary product is contained in $\E{0}$.
\end{theorem}
\begin{theorem} \label{fo-prod-sigma-zero}
	If an extension of $\sqso(\loge{0})$ is contained in $\E{0}$, its grammar does not allow first-order product.
\end{theorem}

For every logic $\LL$, we define an $\LL$-extended quantifier-free (QF) formula as follows:
\begin{eqnarray*}
	\varphi &::=& \alpha, \alpha \text{ is an $\LL$-formula} \ \mid \\
	&& X_i(x_1,\dots,x_{a_i}), i\in\N \ \mid \ \\
	&& (\neg \varphi) \ \mid \ (\varphi \wedge \varphi) \ \mid \ (\varphi \vee \varphi).
\end{eqnarray*}
